Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-07-26 pirouette_supplement.tex
Fix figures, thanks @rsetienne

Based on the 'true alignment' (center-left panel), the candidate model
with the highest marginal likelihood is determined, from a set of 
15 models. The set of models consists of all combinations
of all 4 nucleotides substitution models (JC69, HKY, TN93, GTR),
all 2 clock models (strict and relaxed log-normal) and 2 birth-death
models (Yule and Birth-Death), except the inference model used as the
generative model (JC69, strict clock, Yule). The inference model that had the
highest evidence (as shown in Table \ref{tab:true_evidence}) was 
the inference model with a JC69 nucleotide substitution model,
a relaxed log-normal clock model and a Yule tree model.
The resulting posterior trees are shown in the second panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models.The inference model that had the
highest evidence (as shown in Table \ref{tab:twin_evidence}) was 
the inference model with a JC69 nucleotide substitution model,
a strict clock model and a BD tree model (note: this indeed matches how
the twin tree and twin tree were simulated).
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

To assess if the results of the inference are meaningful one important 
parameter is the Effective Sample Size (ESS). This quantity describes how 
many independent trees are sampled from the posterior distributions. 
For reliable results it is good practice to have at 
least $ESS = 200$ (see 
\begin{sloppypar}
  \url{https://beast.community/ess_tutorial}).
\end{sloppypar}
In the following we present the ESS for the posterior distributions of 
the 4 cases shown in Fig. \ref{fig:example_30_full_pipeline}: 
"true" pipeline with generative model (Table \ref{tab:esses_true_gen}), 
"true" pipeline with best candidate model (Table \ref{tab:esses_true_best}), 
"twin" pipeline with generative model (Table \ref{tab:esses_twin_gen}), 
"twin" pipeline with best candidate model (Table \ref{tab:esses_twin_best}). 
We also report the marginal likelihood (or evidence) data for model selection 
performed both in the "true" (Table \ref{tab:true_evidence}) and 
"twin" pipeline (Table \ref{tab:twin_evidence}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
  \centering
  \resizebox {1.0\columnwidth} {!} {
    \begin{tikzpicture}[
      ->,>=stealth',shorten >=1pt,auto,
      node distance=0.5\textheight, 
      semithick
    ]   
    \tikzstyle{every state}=[]
    \node[state, draw=none] (O) [] {
    };   
    \node[state] (A) [right of = O, rectangle] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30_314/true_tree.png}
    };   
    \node[state] (B) [below of = A, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30_314/true_alignment.png}
    };   
    \node[state] (CG) [below of = B, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_posterior_gen.png}
    };   
    \node[state] (DG) [below of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_error_violin_gen.png}
    };   
    \node[state] (CB) [right of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_posterior_best.png}
    };   
    \node[state] (DB) [below of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/true_error_violin_best.png}
    };   
    \node[state] (AT) [right of = A, rectangle, node distance=0.8\textheight] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30_314/twin_tree.png}
    };   
    \node[state] (BT) [below of = AT, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30_314/twin_alignment.png}
    };   
    \node[state] (CTG) [right of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_posterior_gen.png}
    };   
    \node[state] (DTG) [below of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_error_violin_gen.png}
    };   
    \node[state] (CTB) [right of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_posterior_best.png}
    };   
    \node[state] (DTB) [below of = CTB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30_314/twin_error_violin_best.png}
    };   
    \path 
      (O) edge [anchor = south] node {} (A)
      (A) edge [anchor = south] node {} (B)
      (B) edge [anchor = south] node {} (CG)
      (CG) edge [anchor = south] node {} (DG)
      (B) edge [anchor = south east] node {} (CB)
      (CB) edge [anchor = south] node {} (DB)
      (A) edge [anchor = east] node {} (AT)
      (AT) edge [anchor = south] node {} (BT)
      (BT) edge [anchor = south east] node {} (CTG)
      (CTG) edge [anchor = south] node {} (DTG)
      (BT) edge [anchor = south] node {} (CTB)
      (CTB) edge [anchor = south] node {} (DTB)
    ; 
    \end{tikzpicture}
  }
  \caption{Full pirouette pipeline, including comparison to baseline error. 
    The true tree (top left) is used to simulate an alignment. 
    From this alignment two posterior distributions of trees are created: 
    one using the generative model and another one using the inference model 
    with the highest marginal likelihood. 
    For each distribution of trees, a distribution of errors, 
    measured with the nLTT statistic, 
    between the posterior trees and the main trees is drawn. 
    From the true tree also a twin tree is created (right side of the figure)
    which follows the same pipeline, 
    leading to two additional error distributions to use as baseline errors.}
  \label{fig:example_30_full_pipeline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\input{pirouette_example_30/example_30_314/esses_gen.latex}

@@

\input{pirouette_example_30/example_30_314/esses_best.latex}

@@

\input{pirouette_example_30/example_30_314/esses_twin_gen.latex}

@@

\input{pirouette_example_30/example_30_314/esses_twin_best.latex}

@@

\input{pirouette_example_30/example_30_314/evidence_true.latex}

@@

\input{pirouette_example_30/example_30_314/evidence_twin.latex}

>>>>

Based on the 'true alignment' (center-left panel), the candidate model
with the highest marginal likelihood is determined, from a set of 
15 models. The set of models consists of all combinations
of all 4 nucleotides substitution models (JC69, HKY, TN93, GTR),
all 2 clock models (strict and relaxed log-normal) and 2 birth-death
models (Yule and Birth-Death), except the inference model used as the
generative model (JC69, strict clock, Yule). The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_true}) was 
the inference model with a JC69 nucleotide substitution model,
a relaxed log-normal clock model and a Yule tree model.
The resulting posterior trees are shown in the second panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

Based on the 'twin alignment' (center-right panel), the candidate model
with the highest marginal likelihood is determined, from the same 
set of 15 candidate models.The inference model that had the
highest evidence (as shown in Table \ref{tab:evidence_twin}) was 
the inference model with a JC69 nucleotide substitution model,
a strict clock model and a BD tree model (note: this indeed matches how
the twin tree and twin tree were simulated).
From the 'twin alignment' this best candidate inference model is run.
The resulting posterior trees are shown in the fourth panel
of the third row of posteriors in figure \ref{fig:pipeline}.

@@

To assess if the results of the inference are meaningful one important 
parameter is the Effective Sample Size (ESS). This quantity describes how 
many independent trees are sampled from the posterior distributions. 
For reliable results it is good practice to have at 
least $ESS = 200$ (see 
\begin{sloppypar}
  \url{https://beast.community/ess_tutorial}).
\end{sloppypar}
In the following we present the ESS for the posterior distributions of 
the 4 cases shown in Fig. \ref{fig:example_30_full_pipeline}: 
"true" pipeline with generative model (Table \ref{tab:esses_gen}), 
"true" pipeline with best candidate model (Table \ref{tab:esses_best}), 
"twin" pipeline with generative model (Table \ref{tab:esses_twin_gen}), 
"twin" pipeline with best candidate model (Table \ref{tab:esses_twin_best}). 
We also report the marginal likelihood (or evidence) data for model selection 
performed both in the "true" (Table \ref{tab:evidence_true}) and 
"twin" pipeline (Table \ref{tab:evidence_twin}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[H]
  \centering
  \resizebox {1.0\columnwidth} {!} {
    \begin{tikzpicture}[
      ->,>=stealth',shorten >=1pt,auto,
      node distance=0.5\textheight, 
      semithick
    ]   
    \tikzstyle{every state}=[]
    \node[state, draw=none] (O) [] {
    };   
    \node[state] (A) [right of = O, rectangle] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30/true_tree.png}
    };   
    \node[state] (B) [below of = A, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30/true_alignment.png}
    };   
    \node[state] (CG) [below of = B, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/true_posterior_gen.png}
    };   
    \node[state] (DG) [below of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/true_error_violin_gen.png}
    };   
    \node[state] (CB) [right of = CG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/true_posterior_best.png}
    };   
    \node[state] (DB) [below of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/true_error_violin_best.png}
    };   
    \node[state] (AT) [right of = A, rectangle, node distance=0.8\textheight] {
      \includegraphics[height=0.4\textheight]{pirouette_example_30/example_30/twin_tree.png}
    };   
    \node[state] (BT) [below of = AT, rectangle] {
      \includegraphics[height=0.25\textheight]{pirouette_example_30/example_30/twin_alignment.png}
    };   
    \node[state] (CTG) [right of = CB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/twin_posterior_gen.png}
    };   
    \node[state] (DTG) [below of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/twin_error_violin_gen.png}
    };   
    \node[state] (CTB) [right of = CTG, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/twin_posterior_best.png}
    };   
    \node[state] (DTB) [below of = CTB, rectangle] {
      \includegraphics[height=0.3\textheight]{pirouette_example_30/example_30/twin_error_violin_best.png}
    };   
    \path 
      (O) edge [anchor = south] node {} (A)
      (A) edge [anchor = south] node {} (B)
      (B) edge [anchor = south] node {} (CG)
      (CG) edge [anchor = south] node {} (DG)
      (B) edge [anchor = south east] node {} (CB)
      (CB) edge [anchor = south] node {} (DB)
      (A) edge [anchor = east] node {} (AT)
      (AT) edge [anchor = south] node {} (BT)
      (BT) edge [anchor = south east] node {} (CTG)
      (CTG) edge [anchor = south] node {} (DTG)
      (BT) edge [anchor = south] node {} (CTB)
      (CTB) edge [anchor = south] node {} (DTB)
    ; 
    \end{tikzpicture}
  }
  \caption{Full pirouette pipeline, including comparison to baseline error. 
    The true tree (top left) is used to simulate an alignment. 
    From this alignment two posterior distributions of trees are created: 
    one using the generative model and another one using the inference model 
    with the highest marginal likelihood. 
    For each distribution of trees, a distribution of errors, 
    measured with the nLTT statistic, 
    between the posterior trees and the main trees is drawn. 
    From the true tree also a twin tree is created (right side of the figure)
    which follows the same pipeline, 
    leading to two additional error distributions to use as baseline errors.}
  \label{fig:example_30_full_pipeline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\input{pirouette_example_30/example_30/esses_gen.latex}

@@

\input{pirouette_example_30/example_30/esses_best.latex}

@@

\input{pirouette_example_30/example_30/esses_twin_gen.latex}

@@

\input{pirouette_example_30/example_30/esses_twin_best.latex}

@@

\input{pirouette_example_30/example_30/evidence_true.latex}

@@

\input{pirouette_example_30/example_30/evidence_twin.latex}
